On Logic Embeddings and G\"odel's God

### 2015-03-01 Papers/2014/WADT/WADT-4pages.tex
Fix typos and make minor sentence-level improvements.

\begin{abstract}
  We have applied an elegant and flexible  logic embeddings
  approach to verify and automate a prominent philosophical argument:
  the ontological argument for the existence of God. In our ongoing
  computer-assisted study, higher-order automated reasoning tools have
  made some interesting observations, some of which were unknown so
  far.
\end{abstract}

@@

Logic embeddings provide an elegant means to formalize sophisticated
non-classical logics in classical higher-order logic (HOL, Church's
simple type theory \cite{Church40}). In previous work (cf.~\cite{C35}
and the references therein) the embeddings approach has been
successfully applied to automate object-level and meta-level reasoning
for a range of logics and logic combinations with off-the-shelf HOL
theorem provers. This also includes quantified modal logics (QML)
\cite{J23} and quantified conditional logics (QCL) \cite{C37}.  For
many of the embedded logics few or none automated theorem provers did
exist before. HOL is exploited in this approach to encode the
semantics of the logics to be embedded, for example, Kripke semantics
for QMLs \cite{fitting98} or selection function semantics for QCLs
\cite{Stalnaker68}.

@@

The embeddings approach is flexible, because various modal logics (even
with multiple modalities or a mix of varying/cumulative domain
quantifiers) can be easily supported by stating their characteristic
axioms. Moreover, it is relatively simple to implement, because it
does not require any modification in the source code of the
higher-order prover. A minimal encoding of
second-order modal logic KB in TPTP THF syntax \cite{J22} --- this syntax is
accepted by a range of HOL automated theorem provers (ATPs) --- is
exemplarily provided in Fig.~\ref{fig1}.\footnote{Some Notes on THF, which is a concrete syntax for HOL: \texttt{\$i} and
                   \texttt{\$o} represent the HOL base types $i$ and
                   $o$ (Booleans). \texttt{\$i>\$o} encodes a function
                   (predicate) type.  Predicate
                   application as in $A(X,W)$ is encoded as
                   \texttt{((A@X)@W)} or simply as
                   \texttt{(A@X@W)}, i.e., function/predicate application
                   is represented by \texttt{@}; universal
                   quantification and $\lambda$-abstraction as in
                   $\lambda {A}_{i\rightarrow o} \forall {W_i} (A\,W)$
                   and are represented as in
                   \texttt{\^{}[X:\$i>\$o]:![W:\$i]:(A@W)}; comments begin with
                   \texttt{\%}.} The given set of axioms turns
any TPTP THF compliant HOL-ATP in a reasoning tool for second-order modal logic. A
Henkin-style semantics is thereby assumed for both sides, the side of HOL and
for second-order modal logic.

\begin{figure}[h]
\begin{flushright}
\begin{minipage}{.97\textwidth}
\VerbatimInput[frame=single,%
               commandchars=\\\{\},%
               fontfamily=courier,
               fontseries=b,%
               fontsize=\scriptsize,%
               %rulecolor=green,%
               %fillcolor=yellow,%
               %formatcom=blue,%
               framerule=1pt,%
               framesep=5pt,
               numbersep=5pt,
               numbers=left]%
               {Quantified_KB.ax}
\end{minipage}
\end{flushright}
                \vskip-.5em
               \caption{\label{fig1}HOL encoding of second-order modal
                 logic KB in THF syntax. 
                 Modal formulas are mapped to HOL 
                 predicates (with type \texttt{\$i>\$o}); 
                 type \texttt{\$i} now stands for possible worlds. 
                 The modal connectives $\neg$ (\texttt{mnot}), $\vee$ (\texttt{mor}) and
                 $\square$ (\texttt{mbox}), universal quantification for
                 individuals (\texttt{mall\_ind}) and for sets of 
                 individuals  (\texttt{mall\_indset}) are introduced in 
                 lines 7-18. 
                 Validity of lifted modal formulas is
                 defined in an obvious ways (lines 20-21). Symmetry
                 of accessibility relation $r$ is postulated in lines
                 23-26. Hence, second-order KB is realized here; 
                 for logic K the symmetry axiom can be dropped.
                 % \texttt{?}
                   % is the existential quantifier, and
                   % $\neg,\vee,\wedge, and \Rightarrow$ (mat. impl.)
                   % are written as \texttt{\url{~}}, \texttt{|},
                   % \texttt{\&}, and \texttt{=>}. 
                   % Better formatted and easier readable
                   % presentations of our THF code can easily be
                   % generated with the TPTP tools of
                   % Sutcliffe;
                   % here we did minimize space.
                 }
\end{figure}

In recent work \cite{C40,J28} we have applied the embeddings approach
to verify and automate a philosophical argument that has fascinated
philosophers and theologists for about 1000 years: the ontological
argument for the existence of God~\cite{sobel2004logic}. We have
thereby concentrated on G\"odel's \cite{GoedelNotes}, respectively
Scott's \cite{ScottNotes}, modern version of this argument, which
employs a second-order modal logic (S5), for which, until now, no
theorem provers were available.  In our computer-assisted study of the
argument, the HOL provers LEO-II \cite{C26}, Satallax \cite{Satallax}
and Nitpick \cite{Nitpick} have made some interesting observations,
some of which were unknown so far. This is a landmark result, with
media repercussion in a global scale, and yet it is only a glimpse of
what can be achieved by combining computer science, philosophy and
theology.

We briefly summarize some of these observations: Nitpick confirms that
Scott's axioms are consistent, while LEO-II and Satallax demonstrate that
Gödel's original, slightly different axioms are inconsistent. As far
as we are aware of, this is a new result. As experiments with LEO-II
revealed, the problem lies in a subtle difference in the definitions
of the predicate \textit{essence} (characterizing the essential
properties of an entity) between Gödel and Scott. In several philosophy papers, however, this
difference has been commented as a simple cosmetic change.  Moreover,
instead of using modal logic S5, 
LEO-II and Satallax can prove the final theorem (that is,
$\square \exists x . G(x)$, necessarily there exists God) already for
modal logic KB.  This is highly relevant since several philosophers
have criticized G{\"o}del's argument for the use of logic S5.  Axiom
B (symmetry), however, cannot be dropped, which in turn is confirmed by Nitpick.
LEO-II and Stallax can also show that G{\"o}del's and Scott's axioms
imply what is called the modal collapse: $\phi\supset\Box\phi$. This
formula expresses that contingent truth implies necessary truth (which
can even be interpreted as an argument against free will;
cf.~\cite{sobel2004logic}). The modal collapse is probably the most
fundamental criticism put forward against G{\"o}del's resp. Scott's
versions of the argument. Further theorems that can be shown by LEO-II
and Satallax include flawlessness of God and monotheism.

@@

Ongoing and future work concentrates on the systematic study of
G\"odel's and Scott's proofs. We have also begun to study more recent
variants of the argument
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,bjordal99,fuhrmann05:_exist_notwen,fitting02:_types_tableaus_god,Hajek2002,Hajek2008},
which claim to remedy some fundamental problem of G\"odel's and
Scott's proofs, in particular, the modal collapse.  One interesting
and very encouraging observation from these studies is, that the
argumentation granularity as typically applied in these philosophy
papers obviously already matches very well with the automation capabilities of our
HOL-ATPs. The provides good evidence for the potential relevance of our
approach (not only) wrt. other similar applications in metaphysics.

The long-term goal
is to work out a landscape of the detailled logic parameters (e.g.,
constant vs. varying domains, rigid vs. non-rigid terms, logics KB vs.
S4 vs.  S5, etc.) under which the proposed variants of the modern
ontological argument hold or fail.

There is little related work~\cite{oppenheimera11,rushby13}, and this
focuses solely on the comparably simpler, original ontological
argument by Anselm of Canterbury.  These works do not achieve the
close correspondence between the original formulations and the formal
encodings that can be found in our approach and they do also not reach
the same of degree of proof automation.

Our work attests the maturity of contemporary interactive and
automated deduction tools for HOL and
demonstrates the elegance and practical relevance of the
embeddings-based approach. Most importantly, our work opens new
perspectives towards a computational metaphysics.

>>>>

\begin{abstract}
  We have applied an elegant and flexible  logic embedding
  approach to verify and automate a prominent philosophical argument:
  the ontological argument for the existence of God. In our ongoing
  computer-assisted study, higher-order automated reasoning tools have
  made some interesting observations, some of which were previously unknown.
\end{abstract}

@@

Logic embeddings provide an elegant means to formalize sophisticated
non-classical logics in classical higher-order logic (HOL, Church's
simple type theory \cite{Church40}). In previous work (cf.~\cite{C35}
and the references therein) the embeddings approach has been
successfully applied to automate object-level and meta-level reasoning
for a range of logics and logic combinations with off-the-shelf HOL
theorem provers. This also includes quantified modal logics (QML)
\cite{J23} and quantified conditional logics (QCL) \cite{C37}.  For
many of the embedded logics few or no automated theorem provers did
exist before. HOL is exploited in this approach to encode the
semantics of the logics to be embedded, for example, Kripke semantics
for QMLs \cite{fitting98} or selection function semantics for QCLs
\cite{Stalnaker68}.

@@

The embedding approach is flexible, because various modal logics (even
with multiple modalities or a mix of varying/cumulative domain
quantifiers) can be easily supported by stating their characteristic
axioms. Moreover, it is relatively simple to implement, because it
does not require any modification in the source code of the
higher-order prover. A minimal encoding of
second-order modal logic KB in TPTP THF syntax \cite{J22} --- this syntax is
accepted by a range of HOL automated theorem provers (ATPs) --- is
exemplarily provided in Fig.~\ref{fig1}.\footnote{Some Notes on THF, which is a concrete syntax for HOL: \texttt{\$i} and
                   \texttt{\$o} represent the HOL base types $i$ and
                   $o$ (Booleans). \texttt{\$i>\$o} encodes a function
                   (predicate) type.  Predicate
                   application as in $A(X,W)$ is encoded as
                   \texttt{((A@X)@W)} or simply as
                   \texttt{(A@X@W)}, i.e., function/predicate application
                   is represented by \texttt{@}; universal
                   quantification and $\lambda$-abstraction as in
                   $\lambda {A}_{i\rightarrow o} \forall {W_i} (A\,W)$
                   and are represented as in
                   \texttt{\^{}[X:\$i>\$o]:![W:\$i]:(A@W)}; comments begin with
                   \texttt{\%}.} The given set of axioms turns
any TPTP THF compliant HOL-ATP in a reasoning tool for second-order modal logic. A
Henkin-style semantics is thereby assumed for both logics: HOL and
second-order modal logic.

\begin{figure}[h]
\begin{flushright}
\begin{minipage}{.97\textwidth}
\VerbatimInput[frame=single,%
               commandchars=\\\{\},%
               fontfamily=courier,
               fontseries=b,%
               fontsize=\scriptsize,%
               %rulecolor=green,%
               %fillcolor=yellow,%
               %formatcom=blue,%
               framerule=1pt,%
               framesep=5pt,
               numbersep=5pt,
               numbers=left]%
               {Quantified_KB.ax}
\end{minipage}
\end{flushright}
                \vskip-.5em
               \caption{\label{fig1}HOL encoding of second-order modal
                 logic KB in THF syntax. 
                 Modal formulas are mapped to HOL 
                 predicates (with type \texttt{\$i>\$o}); 
                 type \texttt{\$i} now stands for possible worlds. 
                 The modal connectives $\neg$ (\texttt{mnot}), $\vee$ (\texttt{mor}) and
                 $\square$ (\texttt{mbox}), universal quantification for
                 individuals (\texttt{mall\_ind}) and for sets of 
                 individuals  (\texttt{mall\_indset}) are introduced in 
                 lines 7-18. 
                 Validity of lifted modal formulas is
                 defined in the standard way (lines 20-21). Symmetry
                 of accessibility relation $r$ is postulated in lines
                 23-26. Hence, second-order KB is realized here; 
                 for logic K the symmetry axiom can be dropped.
                 % \texttt{?}
                   % is the existential quantifier, and
                   % $\neg,\vee,\wedge, and \Rightarrow$ (mat. impl.)
                   % are written as \texttt{\url{~}}, \texttt{|},
                   % \texttt{\&}, and \texttt{=>}. 
                   % Better formatted and easier readable
                   % presentations of our THF code can easily be
                   % generated with the TPTP tools of
                   % Sutcliffe;
                   % here we did minimize space.
                 }
\end{figure}

In recent work \cite{C40,J28} we have applied the embedding approach
to verify and automate a philosophical argument that has fascinated
philosophers and theologians for about 1000 years: the ontological
argument for the existence of God~\cite{sobel2004logic}. We have
thereby concentrated on G\"odel's \cite{GoedelNotes} 
modern version of this argument and on
Scott's \cite{ScottNotes} modification, which
employ a second-order modal logic (S5) for which, until now, no
theorem provers were available.  In our computer-assisted study of the
argument, the HOL provers LEO-II \cite{C26}, Satallax \cite{Satallax}
and Nitpick \cite{Nitpick} have made some interesting observations,
some of which were unknown so far. This is a landmark result, with
media repercussion in a global scale, and yet it is only a glimpse of
what can be achieved by combining computer science, philosophy and
theology.

We briefly summarize some of these observations: Nitpick confirms that
Scott's axioms are consistent, while LEO-II and Satallax demonstrate that
Gödel's original, slightly different axioms are inconsistent. As far
as we are aware, this is a new result. As experiments with LEO-II
revealed, the problem lies in a subtle difference in the definitions
of the predicate \textit{essence} (characterizing the essential
properties of an entity) between Gödel and Scott. In several philosophy papers\marginpar{citation needed? weaken the "several" claim?}, however, this
difference has been commented as a simple cosmetic change.  Moreover,
instead of using modal logic S5, 
LEO-II and Satallax can prove the final theorem (that is,
$\square \exists x . G(x)$, necessarily there exists God) already for
modal logic KB.  This is highly relevant since several philosophers
have criticized G{\"o}del's argument for the use of logic S5\marginpar{citation needed? weaken the "several" claim?}.  Axiom
B (symmetry), however, cannot be dropped, which in turn is confirmed by Nitpick.
LEO-II and Satallax can also show that G{\"o}del's and Scott's axioms
imply what is called the modal collapse: $\phi\supset\Box\phi$. This
expresses that contingent truth implies necessary truth (which
can even be interpreted as an argument against free will;
cf.~\cite{sobel2004logic}) and is probably the most
fundamental criticism put forward against G{\"o}del's and Scott's
versions of the argument. Other theorems that can be shown by LEO-II
and Satallax include flawlessness of God and monotheism.

@@

Ongoing and future work concentrates on the systematic study of
G\"odel's and Scott's proofs. We have also begun to study more recent
variants of the argument
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,bjordal99,fuhrmann05:_exist_notwen,fitting02:_types_tableaus_god,Hajek2002,Hajek2008},
which claim to remedy some fundamental problems of G\"odel's and
Scott's proofs, especially the modal collapse.  One interesting
and very encouraging observation from these studies is, that the
argumentation granularity typical of these philosophy
papers is already within reach of the capabilities 
of our higher-order automated theorem provers.
This provides good evidence for the potential 
relevance of the embedding approach (not only) w.r.t. 
other similar applications in metaphysics.

The long-term goal
is to methodically determine the range of logical parameters (e.g.,
constant vs. varying domains, rigid vs. non-rigid terms, logics KB vs.
S4 vs. S5, etc.) under which the proposed variants of the modern
ontological argument hold or fail.

There is little related work~\cite{oppenheimera11,rushby13}, and this
focuses solely on the comparably simpler, original ontological
argument by Anselm of Canterbury.  These works do not achieve the
close correspondence between the original formulations and the formal
encodings that can be found in our approach and they also do not reach
the same degree of proof automation.

Our work attests the maturity of contemporary interactive and
automated deduction tools for HOL and
demonstrates the elegance and practical relevance of the
embeddings-based approach. Most importantly, our work opens new
perspectives towards computational metaphysics.
